Nonlinear Observability for Relative Orbit Determination with Angles-Only Measurements \thanks{This research has been supported in part by NSF under the grants CMMI-1243000 (transferred from 1029551), CMMI-1335008, and CNS-1337722.}

### 2018-06-09 Journal/JAS OD EKF/JAS OD EKF Old/JAS_OD_EKF_rev1.tex
journal added

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%%!PS-Adobe-3.0 EPSF-3.0
%%%BoundingBox: 19 19 221 221
%%%CreationDate: Mon Sep 29 1997
%%%Creator: programmed by hand (JK)
%%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

\usepackage{bm}
\usepackage{amssymb,amsmath,times,subfigure,tabularx,booktabs,colortbl,multirow,threeparttable}
\usepackage{subfigure}
%\usepackage[notref,notcite]{showkeys}  % use this to temporarily show labels
\usepackage[colorlinks=true, pdfstartview=FitV, linkcolor=black, citecolor= black, urlcolor= black]{hyperref}
\usepackage{overcite}
\usepackage{footnpag}			      	% make footnote symbols restart on each page

@@

\begin{abstract}
This paper presents nonlinear observability criteria for the relative orbital dynamics represented by the solutions of the two-body problem. It is assumed that a chief is on a circular orbit with a prescribed orbital radius, and it measures lines-of-sight toward a deputy only. A differential geometric method, based on the Lie derivatives, is used to derive sufficient conditions for observability of the orbital properties of the deputy. It is shown that under certain geometric conditions on the relative configuration between the chief and the deputy, the nonlinear relative motion is observable from angles-only measurements. The second part of this paper presents a quantitative measure of observability for the relative orbits, and it is formulated by generalizing the observability Gramian of linear dynamic systems. An extended Kalman filter is also developed to numerically illustrate the observability of nonlinear relative orbits with angles-only measurements and to show correspondence between the proposed observability measure with filtering errors explicitly.

@@

Space-based surveillance or relative navigation is desirable for many spacecraft missions, such as formation control and rendezvous. Spacecraft maneuvers based only on on-board measurements reduce the total operating cost significantly, and it improves safety against communication interruptions with ground stations.  Relative navigation between spacecraft in close-proximity essentially corresponds to space-based orbit determination.  In particular, vision-based navigation and estimation of relative orbits have received attention recently, since optical sensors have the desirable properties of low cost and minimal maintenance, while providing accurate line-of-sight measurements, or angles from a chief to a deputy.

Relative navigation based on angles-only measurements has been investigated in~\cite{WofGelITAES09,Tom11,PatLovPASFMM12}.  The problem is to determine the relative orbit between a chief spacecraft and a deputy spacecraft by using the line-of-sight between the two objects, assuming that the orbit of the chief is prescribed exactly.  Reference \cite{WofGelITAES09} shows that the relative orbit is unobservable from angles-only measurements when linear relative orbital dynamics are assumed, unless there are thrusting maneuvers. Reference \cite{Tom11} investigates observability by using a relative orbit model linearized in terms of spherical coordinates.  Reference~\cite{PatLovPASFMM12} introduces the concept of partial observability to determine a basis vector representing a family of relative orbits, and an initial orbit determination technique is developed for this method. Linear observability analysis is performed numerically for a particular case in~\cite{YimCraPASMM04}.

All of these results are based on linear relative orbital dynamics. It is straightforward to see that the relative orbit is not observable with angles-only measurements through its linearized dynamics. In other words, there are an infinite number of relative orbits that yield the identical line-of-sight measurements, and the orbital distance between a deputy and a chief cannot be determined by angles-only measurements. As such, it is required to study the nonlinear relative orbital dynamics to determine observability with angles-only measurements.

@@

This paper constructs observability criteria for the nonlinear relative orbital dynamics represented by the solutions of the two-body problem without linearization. Assuming that a chief is on a circular orbit with a prescribed orbital radius, nonlinear equations of motion for the relative orbital motion of a deputy with respect to the chief are presented.  A differential geometric method, based on the Lie derivatives of the line-of-sight from the chief to the deputy~\cite{HerKreITAC77}, is applied to obtain sufficient conditions for observability.  It is shown that under certain geometric conditions on the relative configuration between the chief and the deputy, the nonlinear relative motion is observable from angles-only measurements.

Next, we present a new measure of observability for relative orbit determination with angles-only measurements. This is motivated to understand the degree of observability for the relative orbital dynamics, and to identify the class of relative orbits that have stronger or weaker observability. For a given initial condition of relative orbits, the proposed observability measure determines how much the corresponding relative orbits are easy or difficult to estimate, thereby providing a quantitative measure for relative orbit determination.

@@

In short, the main contributions of this paper are constructing sufficient conditions for the observability of nonlinear relative orbits with angles-only measurements that have not been studied before, and proposing a new quantitative measure of observability.

This paper is organized as follows. The relative orbital dynamics and the orbit determination problem are formulated at Section 2. Observability criteria and a observability measure are presented at Section 3 and Section 4, respectively, which are followed by numerical examples and conclusions.

@@

Define a local-vertical, local-horizontal (LVLH) frame as follows. Its origin is located at the chief satellite. The $x$-axis is along the radial direction from the Earth to the chief, and the $y$-axis is along the velocity vector of the chief. The $z$-axis is normal to the orbital plane, and it is parallel to the angular momentum vector of the chief. The LVLH frame is rotating with the angular velocity of $\mathbf{\omega}=[0,0,n]^T\in\Re^3$, where $n=\sqrt{\frac{\mu}{a^3}}$ is the mean motion of the chief satellite, and $\mu$ denotes the gravitational parameter of the Earth. Note that the velocity of the chief is given by $\mathbf{v}_{chief}=[0,na,0]^T\in\Re^3$. Let the relative position of the deputy satellite with respect to the chief be given by $\mathbf{r}=[x,y,z]^T\in\Re^3$ in the LVLH frame.

@@

Considering that the LVLH frame is rotating with the angular velocity of $\mathbf{\omega}=[0,0,n]^T$, the inertial velocity of the deputy satellite is given by
\begin{align*}
\mathbf{v}_{deputy}
=
\mathbf{v}_{chief} + \dot{\r} + \mathbf{\omega}\times \mathbf{r} = 
\begin{bmatrix}
0 \\ na \\ 0
\end{bmatrix}
+
\begin{bmatrix}
\dot x \\ \dot y \\ \dot z
\end{bmatrix}
+
\begin{bmatrix}
-n y \\ nx \\ 0
\end{bmatrix}
=
\begin{bmatrix}
\dot x - ny \\ \dot y + nx + na \\ \dot z
\end{bmatrix}.
\end{align*}
Therefore, the (normalized) kinetic energy of the deputy satellite is 
\begin{align*}
T = \frac{1}{2}\|\mathbf{v}_{deputy}\|^2 = \frac{1}{2} \braces{(\dot x -ny)^2+(\dot y + nx +na)^2 + \dot z^2}.
\end{align*}
The location of the Earth from the chief is given by $[-a,0,0]^T$ in the LVLH frame. Therefore, the position vector of the deputy from the center of the Earth is given by 
$\r_a=[x+a,y,z]^T\in\Re^3$. The gravitational potential energy is 
\begin{align*}
U = -\frac{\mu}{\sqrt{(x+a)^2 + y^2 + z^2}} = -\frac{\mu}{\|\r_a\|}.
\end{align*}
From the above equations, the Lagrangian of the deputy satellite is expressed as $L(x,y,z,\dot x,\dot y ,\dot z) = T(x,y,\dot x,\dot y,\dot z)- U(x,y,z)$. Using the Euler-Lagrange equations, given by
\begin{align*}
\frac{d}{dt}\deriv{L}{\dot q}-\deriv{L}{q}=0,
\end{align*}
for $q\in\{x,y,z\}$, we obtain the nonlinear equations of motion for the relative orbit as follows.
\begin{align}
\ddot x - 2n\dot y -n^2 x&=n^2 a - \frac{\mu (x+a)}{((x+a)^2 + y^2 + z^2)^{3/2}},\label{eqn:ddotx}\\
\ddot y + 2n\dot x -n^2 y &=   - \frac{\mu y}{((x+a)^2 + y^2 + z^2)^{3/2}},\label{eqn:ddoty}\\
\ddot z &= - \frac{\mu z}{((x+a)^2 + y^2 + z^2)^{3/2}}.\label{eqn:ddotz}
\end{align}

@@

We assume that the line-of-sight from the chief to the deputy is measured by onboard sensors, such as optical sensors. The measurement is represented by the unit-vector along the relative position vector, i.e.,
\begin{align}
\y = h(\mathbf{x}) =\frac{\r}{\|\r\|},\label{eqn:y}
\end{align}
where $\y\in\Re^M$ with $M=3$, and it satisfies $\|\y\|=1$.

@@

Based on the nonlinear dynamic model presented at the previous section, here we derive sufficient conditions for the observability of the relative orbit with line-of-sight measurements.

@@

Observability of nonlinear systems and sufficient conditions for observability are summarized as follows~\cite{HerKreITAC77}. For a given nonlinear dynamic system described with \refeqn{xxdot} and \refeqn{y}, a pair of points $\x_0$ and $\x_1$ is called \textit{indistinguishable} if the outputs of the corresponding solutions starting from each of $\x_0$ and $\x_1$ are identical for a certain time period. Note that $\x_0$ is indistinguishable from $\x_0$ trivially from the definition. The system is \textit{locally weakly observable} at $\x_0$, if there exists an open neighborhood $V$ of $\x_0$ such that for every open neighborhood $V_0$ of $\x_0$ contained in $V$, the only indistinguishable point to $\x_0$ is the point $\x_0$ itself. This implies that $\x_0$ is distinguishable from any other points in a neighborhood of it.

Next, sufficient conditions for observability are presented. The Lie-derivative of the output $h(\x)$ along $f(\x)$, namely $L_f h(\x)$ is defined as follows:
\begin{align*}
L_f h(\x) = \deriv{h(\x)}{\x}f(\x)\in\Re^{M\times 1},
\end{align*}
which corresponds to the directional derivative of $h(\x)$ along $f(\x)$, and $\deriv{h(\x)}{\x}\in\Re^{M\times N}$. For a non-negative integer $i$, the $i$-th order Lie-derivative is defined by induction as $L_f^i h = L_f (L_f^{i-1} h)$ with $L_f^0 h = h$. Define an observability matrix $\mathcal{O}\in\Re^{NM\times N}$ as
\begin{align*}
\mathcal{O}(\x_0) = \deriv{}{\x} \begin{bmatrix} L_f^0 h(\x) \\ L_f^1 h(\x)\\ \vdots \\L_f^{N-1} h(\x)\end{bmatrix}\Bigg|_{\x=\x_0}.
\end{align*}
It has been shown that the system is locally weakly observable at $\x_0$ if the rank of the observability matrix $\mathcal{O}(\x_0) = N$. When applied to linear dynamics, this yields the well-known observability rank condition for linear systems. Note that when there is more than a single measurement, i.e., $M>1$, the observability rank condition can be satisfied without need for computing the higher-order Lie derivatives up to the $N-1$-th order.

@@

\paragraph{Observability Matrix}
We apply the above observability criteria for the nonlinear relative orbital dynamics. The first block of rows of the observability matrix is simply given by
\begin{align*}
\deriv{}{\x} L_f^0 h(\x) = \deriv{\y}{\x}.
\end{align*} 
Similarly, from \refeqn{xxdot}, the second row block is
\begin{align*}
\deriv{}{\x} L_f^1 h(\x) = \deriv{}{\x}\parenth{\deriv{\y}{\x}f(\x)}=\deriv{\dot\y}{\x}.
\end{align*} 
Since $\dot\y$ is determined completely from $\x$, we have
\begin{align*}
\deriv{}{\x} L_f^2 h(\x) = \deriv{}{\x}\parenth{\deriv{\dot\y}{\x}f(\x)}=\deriv{\ddot\y}{\x}.
\end{align*}
Using these and the fact that $\x=[\r^T,\dot \r^T]$, the observability matrix of the relative orbital dynamics can be written as
\begin{align}
\mathcal{O} =
\begin{bmatrix}
\deriv{\y}{\r} & \deriv{\y}{\dot\r}\\
\deriv{\dot\y}{\r} & \deriv{\dot\y}{\dot\r}\\
\deriv{\ddot\y}{\r} & \deriv{\ddot\y}{\dot\r}
\end{bmatrix}
\triangleq
\begin{bmatrix}
\mathcal{O}_{00} & 0_{3\times 3}\\
\mathcal{O}_{10} & \mathcal{O}_{11}\\
\mathcal{O}_{20} & \mathcal{O}_{21}
\end{bmatrix},\label{eqn:OO}
\end{align}
where the submatrices $\mathcal{O}_{ij}\in\Re^{3\times 3}$ are defined for each block, and we used the fact that the measurement $\y$ is independent of $\dot\r$. Here we consider the part of the observability matrix obtained by the second order Lie derivatives of the measurement due to complexity. But, this still provides sufficient conditions for observability with interesting geometric meanings described later.

@@

We wish to show that there is no non-zero vector $\mathbf{c}$ satisfying \refeqn{Oc} under the given conditions. At the first three rows of \refeqn{Oc}, we have
\begin{align}
\mathcal{O}_{00} \mathbf{c}_1 = \frac{1}{\|\r\|}(I-\y\y^T) \mathbf{c}_1 = 0.\label{eqn:c10}
\end{align}
The matrix $\mathcal{O}_{00}$ has a one-dimensional null space spanned by $\y$. Therefore, without loss of generality, we can choose $\mathbf{c}_1=0$ or $\mathbf{c}_1=\r$. As we are interested in any non-zero vector satisfying the above equation, let $\mathbf{c}_1=\r$ for the subsequent development.

@@

Therefore, there is no $\mathbf{c}\in\Re^6$ satisfying \refeqn{Oc} if $\mathbf{c}_1=\r$, under the given condition \refeqn{cond1}. This implies that $\mathbf{c}_1=0$ from \refeqn{c10}. Substituting this back to \refeqn{Oc}, we have $\mathcal{O}_{11}\mathbf{c}_2=\mathcal{O}_{00}\mathbf{c}_2=0$, which follows that $\mathbf{c}_2=c\r$ for some constant $c$. However, when $\mathbf{c}_2=c\r$, the last three rows of \refeqn{Oc} are given by
\begin{align}
\mathcal{O}_{21}\mathbf{c}_2 = 2c(\mathcal{O}_{10}-\mathcal{O}_{00}[\omega]_\times) \r
= -2c\mathcal{O}_{00}\mathbf{v}_{rel}=0,\label{eqn:row3i1}
\end{align}
where \refeqn{O10r} is used. But, from the first condition of \refeqn{cond1}, we have $\mathcal{O}_{00}\mathbf{v}_{rel}\neq 0$, and therefore $c=0$, i.e., $\mathbf{c}_2=c\r =0$.

@@

\paragraph{Case (ii): $\r\times\dot \r\neq0$} Next, we consider the second case of \refeqn{row2}. It implies that $-\dot\r + \mathbf{c}_2$ is parallel to $\r$, or equivalently, $\mathbf{c}_2 = \dot\r + c\r$ for an arbitrary constant $c$. For the given choice of $\mathbf{c}=[\r^T,\dot\r+ c\r^T]^T$, the last three rows of \refeqn{Oc} are given by
\begin{align*}
\mathcal{O}_{20} \mathbf{c}_1 +\mathcal{O}_{21} \mathbf{c}_2 
& = (\mathcal{O}_{20} \r+\mathcal{O}_{21}\dot\r) + c\mathcal{O}_{21} \r.
\end{align*}
From \refeqn{O20r}, \refeqn{O21r}, \refeqn{O21rdot}, this can be rewritten as
\begin{align*}
\mathcal{O}_{20} \mathbf{c}_1 +\mathcal{O}_{21} \mathbf{c}_2 
= -\mathcal{O}_{00}(\mathbf{a}_2+2c\mathrm{v}_{ref})= -\frac{1}{\|\r\|}(I-\y\y^T)(\mathbf{a}_2+2c\mathbf{v}_{rel})=0.
\end{align*}
By following the same argument given after \refeqn{row3i}, under \refeqn{cond2}, there is no $c$ satisfying the above equation.

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required

%\begin{filecontents*}{example.eps}
%%%!PS-Adobe-3.0 EPSF-3.0
%%%%BoundingBox: 19 19 221 221
%%%%CreationDate: Mon Sep 29 1997
%%%%Creator: programmed by hand (JK)
%%%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}

%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

\usepackage{bm}
\usepackage{amssymb,amsmath,times,subfigure,tabularx,booktabs,colortbl,multirow,threeparttable}
\usepackage{subfigure}
%\usepackage[notref,notcite]{showkeys}  % use this to temporarily show labels
\usepackage[colorlinks=true, pdfstartview=FitV, linkcolor=black, citecolor= black, urlcolor= black]{hyperref}
%\usepackage{overcite}
%\usepackage{footnpag}			      	% make footnote symbols restart on each page

@@

\begin{abstract}
This paper presents nonlinear observability criteria for the relative orbital dynamics represented by the solutions of the two-body problem. It is assumed that a chief is on a circular orbit with a prescribed orbital radius, and it measures lines-of-sight toward a deputy only. A differential geometric method, based on the Lie derivatives, is used to derive sufficient conditions for observability of the orbital properties of the deputy. It is shown that under certain geometric conditions on the relative configuration between the chief and the deputy, the nonlinear relative motion is observable from angles-only measurements. The second part of this paper presents a quantitative measure of observability for the relative orbits, and it is formulated by generalizing the observability Gramian of linear dynamic systems. An extended Kalman filter is also developed to numerically illustrate the observability of nonlinear relative orbits with angles-only measurements and to show correspondence between the proposed observability measure and filtered solution accuracy.

@@

%Space-based surveillance or relative navigation is desirable for many spacecraft missions, such as formation control and rendezvous. Spacecraft maneuvers based only on on-board measurements reduce the total operating cost significantly, and it improves safety against communication interruptions with ground stations. Relative navigation between spacecraft in close-proximity essentially corresponds to space-based orbit determination. In particular, vision-based navigation and estimation of relative orbits have received attention recently, since optical sensors have the desirable properties of low cost and minimal maintenance, while providing accurate line-of-sight measurements, or angles from a chief to a deputy.

Relative orbit determination (ROD) between spacecraft is desirable for many spacecraft missions. In scenarios where the spacecraft are in close-proximity, this is commonly referred to as relative navigation.
When baselines between the spacecraft are longer, a more appropriate term would be space-based space surveillance or space-based orbit determination. In particular, vision-based ROD has received attention recently, since optical sensors have the desirable properties of being low cost and minimal maintenance, while providing accurate line-of-sight measurements, or angles, from a chief to a deputy.

ROD based on angles-only measurements has been investigated in \cite{WofGelITAES09,PatLovPASFMM12}. The problem is to determine the relative orbit between a chief spacecraft and a deputy spacecraft by using the line-of-sight between the two objects, assuming that the orbit of the chief is prescribed exactly. Reference \cite{WofGelITAES09} shows that the relative orbit is unobservable from angles-only measurements when linear relative orbital dynamics are assumed, unless there are thrusting maneuvers. Reference~\cite{PatLovPASFMM12} introduces the concept of partial observability to determine a basis vector representing a family of relative orbits, and an initial orbit determination technique is developed for this method. Linear observability analysis is performed numerically for a particular case in~\cite{YimCraPASMM04}.

All of the above results are based on linear relative orbital dynamics.
%It is straightforward to see that the relative orbit is not observable with angles-only measurements through its linearized dynamics. In other words,
Under these assumptions, there are an infinite number of relative orbits that yield identical line-of-sight measurements, and therefore the orbital distance between a deputy and a chief cannot be determined by angles-only measurements. As such, it is required to study the nonlinear relative orbital dynamics to determine observability with angles-only measurements.

@@

This paper constructs observability criteria for the nonlinear relative orbital dynamics represented by the two-body problem without linearization. Assuming that a chief is on a circular orbit with a prescribed orbital radius, nonlinear equations of motion for the relative orbital motion of a deputy with respect to the chief are presented. A differential geometric method, based on the Lie derivatives of the line-of-sight from the chief to the deputy~\cite{HerKreITAC77}, is applied to obtain sufficient conditions for observability, under the assumptions of these equations. It is shown that under certain geometric conditions on the relative configuration between the chief and the deputy, the relative motion is observable from angles-only measurements if nonlinear dynamics are assumed.

Next, we present a new measure of observability for relative orbit determination with angles-only measurements. The objectives are to understand the degree of observability for the relative orbital dynamics, and to identify the class of relative orbits that have stronger or weaker observability. For a given initial condition of relative orbits, the proposed observability measure determines how much the corresponding relative orbits are easy or difficult to estimate, thereby providing a quantitative measure for relative orbit determination.

@@

In short, the main contributions of this paper are constructing sufficient conditions for the observability of nonlinear relative orbits with angles-only measurements, and proposing a new quantitative measure of observability. The main objective is to provide assurance that relative orbits are observable with angles-only measurements under certain conditions, if the nonlinearity of relative orbital dynamics are properly considered. More comprehensive study for observability of nonlinear relative orbits, such as obtaining necessary conditions for observability or sufficient conditions for unobservability, is deferred for future study.

This paper is organized as follows. The relative orbital dynamics and the orbit determination problem are formulated in Section 2. Observability criteria and an observability measure are presented in Section 3 and Section 4, respectively, which are followed by numerical examples and conclusions.

@@

Define a local-vertical, local-horizontal (LVLH) frame as follows. Its origin is located at the chief satellite. The $x$-axis is along the radial direction from the Earth to the chief, and the $y$-axis is along the velocity vector of the chief. The $z$-axis is normal to the orbital plane, and it is parallel to the angular momentum vector of the chief. The LVLH frame is rotating with the angular velocity of $\mathbf{\omega}=[0,0,n]^T\in\Re^3$, where $n=\sqrt{\frac{\mu}{a^3}}$ is the mean motion of the chief satellite, and $\mu$ denotes the gravitational parameter of the Earth. Note that the inertial velocity of the chief expressed in LVLH coordinates is given by $\mathbf{v}_{chief}=[0,na,0]^T\in\Re^3$. Let the relative position of the deputy satellite with respect to the chief be given by $\mathbf{r}=[x,y,z]^T\in\Re^3$ in the LVLH frame.

@@

Considering that the LVLH frame is rotating with the angular velocity of $\mathbf{\omega}=[0,0,n]^T$, the inertial velocity of the deputy satellite expressed in LVLH coordinates is given by
\begin{align*}
\mathbf{v}_{deputy}
=
\mathbf{v}_{chief} + \dot{\r} + \mathbf{\omega}\times \mathbf{r} = 
\begin{bmatrix}
0 \\ na \\ 0
\end{bmatrix}
+
\begin{bmatrix}
\dot x \\ \dot y \\ \dot z
\end{bmatrix}
+
\begin{bmatrix}
-n y \\ nx \\ 0
\end{bmatrix}
=
\begin{bmatrix}
\dot x - ny \\ \dot y + nx + na \\ \dot z
\end{bmatrix}.
\end{align*}
Therefore, the (normalized) kinetic energy of the deputy satellite is 
\begin{align*}
T = \frac{1}{2}\|\mathbf{v}_{deputy}\|^2 = \frac{1}{2} \braces{(\dot x -ny)^2+(\dot y + nx +na)^2 + \dot z^2}.
\end{align*}
The location of the Earth from the chief is given by $[-a,0,0]^T$ in the LVLH frame. Therefore, the position vector of the deputy from the center of the Earth is given by 
$\r_a=[x+a,y,z]^T\in\Re^3$. The gravitational potential energy is 
\begin{align*}
U = -\frac{\mu}{\sqrt{(x+a)^2 + y^2 + z^2}} = -\frac{\mu}{\|\r_a\|}.
\end{align*}
From the above equations, the Lagrangian of the deputy satellite is expressed as $L(x,y,z,\dot x,\dot y ,\dot z) = T(x,y,\dot x,\dot y,\dot z)- U(x,y,z)$. Using the Euler-Lagrange equations, given by
\begin{align*}
\frac{d}{dt}\deriv{L}{\dot q}-\deriv{L}{q}=0,
\end{align*}
for $q\in\{x,y,z\}$, we obtain the nonlinear equations of motion for the relative orbit as follows.
\begin{align}
\ddot x - 2n\dot y -n^2 x&=n^2 a - \frac{\mu (x+a)}{((x+a)^2 + y^2 + z^2)^{3/2}},\label{eqn:ddotx}\\
\ddot y + 2n\dot x -n^2 y &=   - \frac{\mu y}{((x+a)^2 + y^2 + z^2)^{3/2}},\label{eqn:ddoty}\\
\ddot z &= - \frac{\mu z}{((x+a)^2 + y^2 + z^2)^{3/2}}.\label{eqn:ddotz}
\end{align}

@@

We assume that the line-of-sight from the chief to the deputy is measured by onboard sensors, such as optical sensors. The measurement is represented by the unit-vector along the relative position vector, i.e.,
\begin{align}
\y = h(\mathbf{x}) =\frac{\r}{\|\r\|},\label{eqn:y}
\end{align}
where $\y\in\Re^3$, and it satisfies $\|\y\|=1$.

@@

Based on the nonlinear dynamic model presented in the previous section, here we derive sufficient conditions for the observability of the relative orbit with line-of-sight measurements.

@@

Observability of nonlinear systems and sufficient conditions for observability are summarized as follows~\cite{HerKreITAC77,Nijvan90}. For a given nonlinear dynamic system described with \refeqn{xxdot} and \refeqn{y}, a pair of points $\x_0$ and $\x_1$ is called \textit{indistinguishable} if the outputs of the corresponding solutions starting from each of $\x_0$ and $\x_1$ are identical for a certain time period. Note that $\x_0$ is indistinguishable from $\x_0$ trivially from the definition. The system is \textit{locally weakly observable} at $\x_0$, if there exists an open neighborhood $V$ of $\x_0$ such that for every open neighborhood $V_0$ of $\x_0$ contained in $V$, the only indistinguishable point to $\x_0$ is the point $\x_0$ itself. This implies that $\x_0$ is distinguishable from any other points in a certain neighborhood of it.

Next, sufficient conditions for observability are presented. The Lie-derivative of the output $h(\x)$ along $f(\x)$, namely $L_f h(\x)$ is defined as follows:
\begin{align*}
L_f h(\x) = \deriv{h(\x)}{\x}f(\x)\in\Re^{M\times 1},
\end{align*}
which corresponds to the directional derivative of $h(\x)$ along $f(\x)$, and $\deriv{h(\x)}{\x}\in\Re^{M\times N}$. For a non-negative integer $i$, the $i$-th order Lie-derivative is defined by induction as $L_f^i h = L_f (L_f^{i-1} h)$ with $L_f^0 h = h$. For a positive integer $k$, define a $k$-th order observability matrix $\mathcal{O}\in\Re^{kM\times N}$ as
\begin{align*}
\mathcal{O}(\x_0) = \deriv{}{\x} \begin{bmatrix} L_f^0 h(\x) \\ L_f^1 h(\x)\\ \vdots \\L_f^{k} h(\x)\end{bmatrix}\Bigg|_{\x=\x_0}.
\end{align*}
It has been shown that the system is locally weakly observable at $\x_0$ if the rank of the observability matrix $\mathcal{O}(\x_0) = N$ for some $k$~\cite{HerKreITAC77,Nijvan90}. When applied to linear dynamics, this yields the well-known observability rank condition for linear systems.

In general, the order $k$ of the observability should be increased until the rank condition is satisfied. Note that when there is more than a single measurement type for each measurement time, i.e., $M>1$, it is possible to satisfy the observability rank condition without need for computing the higher-order Lie derivatives up to the $N-1$-th order.

@@

\paragraph{Observability Matrix}
We apply the above observability criteria for the nonlinear relative orbital dynamics. The first block of rows of the observability matrix is simply given by
\begin{align*}
\deriv{}{\x} L_f^0 h(\x) = \deriv{\y}{\x}.
\end{align*} 
Similarly, from \refeqn{xxdot}, the second row block is
\begin{align*}
\deriv{}{\x} L_f^1 h(\x) = \deriv{}{\x}\parenth{\deriv{\y}{\x}f(\x)}=\deriv{\dot\y}{\x}.
\end{align*} 
Since $\dot\y$ is determined completely from $\x$, we have
\begin{align*}
\deriv{}{\x} L_f^2 h(\x) = \deriv{}{\x}\parenth{\deriv{\dot\y}{\x}f(\x)}=\deriv{\ddot\y}{\x}.
\end{align*}
Using these and the fact that $\x=[\r^T,\dot \r^T]$, the observability matrix of the relative orbital dynamics can be written as
\begin{align}
\mathcal{O} =
\begin{bmatrix}
\deriv{\y}{\r} & \deriv{\y}{\dot\r}\\[0.4em]
\deriv{\dot\y}{\r} & \deriv{\dot\y}{\dot\r}\\[0.4em]
\deriv{\ddot\y}{\r} & \deriv{\ddot\y}{\dot\r}
\end{bmatrix}
\triangleq
\begin{bmatrix}
\mathcal{O}_{00} & 0_{3\times 3}\\
\mathcal{O}_{10} & \mathcal{O}_{11}\\
\mathcal{O}_{20} & \mathcal{O}_{21}
\end{bmatrix}\in\Re^{9\times 6},\label{eqn:OO}
\end{align}
where the submatrices $\mathcal{O}_{ij}\in\Re^{3\times 3}$ are defined for each block, and we used the fact that the measurement $\y$ is independent of $\dot\r$. Here we consider the observability matrix obtained by up to the second order Lie derivatives of the measurement due to complexity. But, this still provides sufficient conditions for observability under certain geometric conditions described later.

@@

We wish to show that there is no non-zero vector $\mathbf{c}$ satisfying \refeqn{Oc} under the given conditions. In the first three rows of \refeqn{Oc}, we have
\begin{align}
\mathcal{O}_{00} \mathbf{c}_1 = \frac{1}{\|\r\|}(I-\y\y^T) \mathbf{c}_1 = 0.\label{eqn:c10}
\end{align}
The matrix $\mathcal{O}_{00}$ has a one-dimensional null space spanned by $\y$. Therefore, without loss of generality, we can choose $\mathbf{c}_1=0$ or $\mathbf{c}_1=\r$. As we are interested in any non-zero vector satisfying the above equation, let $\mathbf{c}_1=\r$ for the subsequent development.

@@

Therefore, there is no $\mathbf{c}\in\Re^6$ satisfying \refeqn{Oc} if $\mathbf{c}_1=\r$, under the given condition \refeqn{cond1}. This implies that $\mathbf{c}_1=0$ from \refeqn{c10}. Substituting this back into \refeqn{Oc}, we have $\mathcal{O}_{11}\mathbf{c}_2=\mathcal{O}_{00}\mathbf{c}_2=0$, which follows that $\mathbf{c}_2=c\r$ for some constant $c$. However, when $\mathbf{c}_2=c\r$, the last three rows of \refeqn{Oc} are given by
\begin{align}
\mathcal{O}_{21}\mathbf{c}_2 = 2c(\mathcal{O}_{10}-\mathcal{O}_{00}[\omega]_\times) \r
= -2c\mathcal{O}_{00}\mathbf{v}_{rel}=0,\label{eqn:row3i1}
\end{align}
where \refeqn{O10r} is used. But, from the first condition of \refeqn{cond1}, we have $\mathcal{O}_{00}\mathbf{v}_{rel}\neq 0$, and therefore $c=0$, i.e., $\mathbf{c}_2=c\r =0$.

@@

\paragraph{Case (ii): $\r\times\dot \r\neq0$} Next, we consider the second case of \refeqn{row2}. It implies that $-\dot\r + \mathbf{c}_2$ is parallel to $\r$, or equivalently, $\mathbf{c}_2 = \dot\r + c\r$ for an arbitrary constant $c$. For the given choice of $\mathbf{c}=[\r^T,\dot\r+ c\r^T]^T$, the last three rows of \refeqn{Oc} are given by
\begin{align*}
\mathcal{O}_{20} \mathbf{c}_1 +\mathcal{O}_{21} \mathbf{c}_2 
& = (\mathcal{O}_{20} \r+\mathcal{O}_{21}\dot\r) + c\mathcal{O}_{21} \r.
\end{align*}
From \refeqn{O20r}, \refeqn{O21r}, \refeqn{O21rdot}, this can be rewritten as
\begin{align*}
\mathcal{O}_{20} \mathbf{c}_1 +\mathcal{O}_{21} \mathbf{c}_2 
= -\mathcal{O}_{00}(\mathbf{a}_2+2c\mathbf{v}_{ref})= -\frac{1}{\|\r\|}(I-\y\y^T)(\mathbf{a}_2+2c\mathbf{v}_{rel})=0.
\end{align*}
By following the same argument given after \refeqn{row3i}, under \refeqn{cond2}, there is no $c$ satisfying the above equation.

This implies that $\mathbf{c}_1=0$. Then, by the same argument given in \refeqn{row3i1}, we have $\mathbf{c}_2=0$. In short, under the given condition \refeqn{cond2}, the equation \refeqn{Oc} implies that $\mathbf{c}=0$. Therefore, the six columns of the observability matrix $\mathcal{O}$ are linearly independent. 
\end{proof}
